---
title: "RCT Analysis Report"
author: "Dr. Sean Sylvia"
date: today
format:
  html: default
  pdf: default
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

# Introduction

This report analyzes a randomized controlled trial (RCT) dataset using R. The goal is to demonstrate reproducible data analysis practices, including data loading, cleaning, and statistical analysis.

---

## Load Packages and Functions
```{r}
# Load necessary packages and custom helper functions
library(tidyverse)
library(broom)
library(renv)
source("helpers.R")
```

---

## Load Data

```{r}
# Load the raw dataset
data <- read_csv("../data/raw/rct_sim_data.csv")

# Preview the dataset
head(data)
```

---

## Data Cleaning

```{r}
# Clean the data using a helper function
cleaned_data <- clean_rct_data(data)

# Check for missing values
summary(cleaned_data)
```

---

## Descriptive Statistics

```{r}
# Summarize the dataset using a helper function
summarize_data(cleaned_data)
```

---

## Covariate Balance

```{r}
# Assess balance of covariates between treatment and control groups
covariate_balance <- check_covariate_balance(cleaned_data)
covariate_balance
```

---

## Difference-in-Means Analysis

```{r}
# Run a basic difference-in-means analysis
ate_estimate <- difference_in_means(cleaned_data)
ate_estimate
```

---

## Plot Treatment Effects

```{r}
# Generate a simple plot of treatment effects
plot_treatment_effects(cleaned_data)
```

